Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-10 paper.tex
Updates in II.A and II.B

Split and merge relate to the same dual problem in our context. We
focus here on merging for the sake of concision. Splitting algorithms
can be derived from merging ones by swapping reads and writes. Our
goal then is to merge a set of $n$ chunks into a single reconstructed
3D image with $R$ voxels of size $b$. For simplicity, we assume that
all chunks are of identical size, that they do not overlap, and that
slices are squares and blocks are cubes.

We consider a file format where voxels are written in column-major
order. All voxels in a \emph{slice} have a the same $k$ and all voxels in
a \emph{row} have the same $j$.  NIfTI is such a format,
used by the vast majority of the neuroimaging community.

@@

We adopt the following notations (see Figure\ref{fig:notations}):
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available memory (in B).
\item $m'$: amount of used memory (in B), $m'\leq m$.
\end{itemize}
We also have the following relations:
\begin{itemize}
\item Number of slices, rows, columns in a block: $\sqrt[3]{\frac{R}{n}}=d$.
\item Number of blocks in a block row: $\sqrt[3]{n}$.
\end{itemize}

@@

A disk is characterized by its read and write rates and its seek
time. We do not distinguish between access time and seek time. We also
assume that seeks require a constant amount of time, regardless of the
position seeked to. In practice, large variations would be expected,
but modeling such variations would inevitably lead to models specific
to the hardware, file system or operating system, which we
intentionally avoid here. Likewise, in modern systems, read and write
times are greatly impacted by caching occurring at several levels,
which we do not model here. Thus, our goal is to find algorithms that
minimize the \emph{number} of seek operations.

\todo{Add ballpark values of seek time (5ms) and read/write rates.}

>>>>

Split and merge relate to the same dual problem in our context. We
focus here on merging for the sake of concision. Splitting algorithms
can be derived from merging ones by swapping reads and writes. Our
goal then is to merge a set of $n$ chunks into a single reconstructed
3D image with $R$ voxels of size $b$. For simplicity, we assume that
chunks are non-overlapping cuboids that all have the same dimension.

Although our algorithms could be applied to any byte organization, we
consider a file format where voxels are written in column-major
order. All voxels in a \emph{slice} have the same $k$ and all voxels
in a \emph{row} have the same $j$.

@@

We adopt the following notations (see Figure~\ref{fig:notations}):
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available memory (in B).
\item $m'<m$: amount of used memory (in B).
\end{itemize}
We also have the following relations:
\begin{itemize}
\item Number of slices/rows/columns in a block: $\sqrt[3]{\frac{R}{n}}=d$.
\item Number of blocks in a block row: $\sqrt[3]{n}$.
\end{itemize}

@@

A disk is characterized by its read and write rates, its access time
and its seek time. For common file sizes, seek time is negligible
compared to read or write time as typical seek times range from about 0.1~ms
for Solid-State Drives (SSD) to 10~ms for Hard-Disk
Drives (HDD). However, as we will shown later, naive algorithms might
seek up to $10^7$ times to merge a high-resolution image, which
renders seek time comparable to read and write times. In addition,
extensive seeking also has an effect on read and write rates, as these
are typically increasing with the duration of uninterrupted reads or
writes.

In our complexity analysis, we do not distinguish between access time
and seek time.  We also assume that seeks require a constant amount of
time, regardless of the position seeked to. That is, we focus on the
average seek time. In practice, large variations would be expected
depending on the seek distance, but modeling such variations would
inevitably lead to models specific to the hardware, file system or
operating system, which we intentionally avoid here. Likewise, in
contemporary systems, read and write times are greatly impacted by caches
operating at several levels, which we do not model here. Thus, our
goal is to find algorithms that minimize the \emph{number} of seek and
file access operations, which we denote ``number of seeks'' in the
remainder.
